Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-06 Manuscript/manuscript.Rmd
intro edits

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and regions that include TAD boundaries. However, <!--this is a multi-faceted probelem. [??? "WATER"!]-->many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. [??? You confused me a lot with this] Such heavily imbalanced datasets <!--are known to [??? WATER!] -->adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how the increased resolution of Hi-C data affect the performance of predictive modeling.

@@

Recently, an L1-regularized multiple logistic regression model was developed to predict TAD boundaries using transcription factor binding sites on GM12878 [@mourad2016computational]. Based on a 1 kb binned genomic data matrix and 20 kb flanked TAD boundaries, the model was trained to identify the set of transcription factors most influential of TAD boundaries. Several problems exist with the study. First, the class imbalance issue was not addressed. Secondly, many transcription factors are known to colocalize, and therefore exhibit high correlations. If a group of predictors are highly correlated among themselves, L1-regularization tends to pick only one of them and will shrink the others to zero [@zou2005regularization].  Thirdly, [@mourad2016computational] reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. It has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role].

>>>>

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and regions that include TAD boundaries. However, <!--this is a multi-faceted probelem. [??? "WATER"!]-->many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. <!--including sparsity of TAD boundaries as well as suitably describing the spatial relationship between annotations and TAD boundaries.-->Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. The imbalance ratio is defined as the ratio of minority classes (regions with TAD boundaries) to majority classes (regions without TAD boundaries) [@Orriols-Puig2008]. [??? You confused me a lot with this] Such heavily imbalanced datasets <!--are known to [??? WATER!] -->adversely impact the performance of the classifiers as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, was developed to predict TAD boundaries using transcription factor binding sites on GM12878 [@mourad2016computational]. Based on a 1 kb binned genomic data matrix and 20 kb flanked TAD boundaries, the model was trained using proportions of overlaps between ChiP-seq peak defined functional element coordinates and genomic bins. However, several problems exist with this study. First, many transcription factors are known to colocalize, and therefore exhibit high correlations. Given a group of features that are highly correlated, L1-regularization tends to pick only one of them, while shrinking the others to zero [@zou2005regularization]. Here, ensemble learning techniques could offer a more robust method toward identifying  annotations associated with TAD boudaries. Likewise, it is unclear how other sets of epigenomic features affect prediction, including chromatin state segmentations and histone modifications. Secondly, the class imbalance issue was not addressed prior to modelling. Instead, model performance was evaluated using ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve). It has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role]. It is unclear how simple data level solutions to class imbalance, such random under-sampling and random over-sampling affect TAD boundary prediction. Likewise, a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance. Thirdly, the relationship between annotations and boundaries was quantified using proportions of overlaps. This type of predictor ignores distances, which offers a much more spatially descriptive feature space and could improve predictive performance. Lastly, it is unclear how *HiCfeat* would perform on other cell lines and Hi-C data resolutions, giving *HiCfeat* a limited application in TAD boundary prediction.

<!--
[Paragraph about disconnect between Hi-C and epigenomic data] Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).
-->

In this study, we developed a computational framework aimed at determining the optimal set of data level characteristics necessary for optimal prediction of TAD boundaries and identification of influential genomic annotations. We utilize random forest classification models built on HiC-data from two cell lines, including GM12878 and K562, obtained at four different resolutions, including 10 kb, 25 kb, 50 kb, and 100 kb. For each cell line and resolution, we compared three feature engineering techniques, including counts, percents, and distances. Additionally, we compared three re-sampling techniques, including random under-sampling, random over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE). Model performances were evaluated and compared using Matthews Correlation Coefficient (MCC).

[???Include concluding paragraph]
